---
title: "Adding a new client brand to hdatools"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 8,
                      dev = "ragg_png")

```

hdatools' theming is built on one idea: **brands are data**. Every
`theme_*()`, `scale_color_*()`/`scale_fill_*()`, `scale_*_*_c()`/`scale_*_*_b()`,
and `*_colors`/`*_color()` export is a thin, brand-agnostic wrapper over a
single internal registry, `.brands` (see `R/brands.R`). Adding a new
client brand means adding **one entry** to that registry — the wrappers
below it are one-liners you copy-and-rename, not new logic.

This article walks through how VHA was added as hdatools' fourth brand,
as the worked example.

## 1. Confirm the brand assets first

Before writing any code, get three things confirmed directly (never invent
or infer them):

- The brand's hex palette, with a label for each color.
- Rights to bundle the brand's font(s) — hdatools ships static `.ttf` files
  in `inst/fonts/`, registered offline via `systemfonts::register_font()`
  (from the bundled file, never fetched from a font server), so redistribution
  rights matter, not just "the font looks free."
- Any design decisions specific to the brand (base text size, output-format
  size adjustments, line height, title/subtitle margins) if the client
  already has an existing house style to match.

VHA's palette and Montserrat's OFL licensing were confirmed in chat before
`R/brands.R` was touched.

## 2. Add the registry entry

Every brand entry in `.brands` has the same shape:

```r
vha = list(
  palette = c(
    "Dark Turq"   = "#0C4D4F",
    "Light Green" = "#A0D18E",
    "Yellow"      = "#ECC51E",
    "Light Turq"  = "#19787B",
    "Grey"        = "#2E3030",
    "Light Blue"  = "#E3F3F5"
  ),
  gradient = NULL,          # only populated for brands with a legacy
                            # scale_*_gradient_*() to soft-deprecate
  na_color = "#d6dbdb",
  fonts = list(title = "Montserrat", body = "Montserrat"),
  base_size = 13,
  html_adjust = 4,
  pdf_adjust = 7,
  lineheight = 0.9,
  theme_fonts = list(
    title = "Montserrat", subtitle = "Montserrat",
    caption = "Montserrat", strip = NULL   # NULL tracks the caller's base_family
  ),
  theme_margins = list(
    title = ggplot2::margin(b = 10, unit = "pt"),
    subtitle = ggplot2::margin(t = -5, b = 10, unit = "pt")
  ),
  ramps = list(
    sequential = list(h1 = 196.3, c1 = 22, l1 = 27, l2 = 95, power = 0.90),
    diverging  = list(h1 = 196.3, h2 = 67.6, c1 = 70, l1 = 40, l2 = 95, power = 0.90)
  )
)
```

The `ramps` HCL parameters aren't hand-picked colors — `h1`/`h2` are the
sequential/diverging anchor colors' own HCL hue (compute with
`colorspace::hex2RGB(hex) |> methods::as("polarLUV")`); `c1`/`l1` follow that
anchor's chroma/lightness; `l2 = 95` and the cream light-end hue/chroma are
fixed constants shared by every brand (`R/ramps.R`). Render a swatch and run
a CVD simulation (`colorspace::deutan()`/`protan()`/`tritan()`) before
committing the params — a hue whose natural lightness is far from the dark
anchor (VHA's Yellow, used for the diverging ramp's second arm) can render
muddy or lose its identity at the anchor lightness. That's an sRGB gamut
limit, not a mistake; document it as a provisional ramp (see `scale_color_vha_c()`)
rather than silently shipping a ramp that looks wrong.

If the brand introduces a new font family, bundle its static `.ttf` files
under `inst/fonts/<family>/` alongside its license file, add a manifest row
to `inst/fonts/LICENSES.md`, and register it in `register_hda_fonts()`
(`R/theme_helpers.R`) with `systemfonts::register_font()`.

## 3. Add the one-line wrapper exports

Every other brand-specific symbol is a direct copy of an existing brand's
wrapper, renamed:

- **`R/colors.R`**: `vha_colors <- .brands$vha$palette` and
  `vha_color <- .brand_color(.brands$vha$palette, "VHA")`.
- **`R/themes.R`**: `theme_vha()` calls `.brand_theme(.brands$vha, brand = "vha", ...)`
  with the brand's own `base_size`/`base_family`/`html_adjust`/`pdf_adjust`
  defaults.
- **`R/scales.R`**: `scale_color_vha()`/`scale_fill_vha()` (+ `scale_colour_vha()`
  alias) call `.scale_brand_discrete()`; `scale_*_vha_c()`/`scale_*_vha_b()`
  (+ aliases) call `.scale_brand_continuous()`/`.scale_brand_binned()`.

None of `R/scales.R`'s or `R/themes.R`'s shared internals (`.brand_theme()`,
`.scale_brand_discrete()`, `.scale_brand_continuous()`, `.scale_brand_binned()`,
`.ramp_hex()`) needed to change, or even know VHA exists — they all take
`brand` as a string and read everything else from `.brands[[brand]]`. If
adding a brand ever requires an `if (brand == "...")` branch in one of those
internals, that's a sign the shared builder isn't general enough yet — worth
flagging before hand-rolling a brand-specific exception.

## 4. Document and test

Run `devtools::document()` to pick up the new roxygen blocks (`NAMESPACE`
and `man/*.Rd` are generated — never hand-edit them). Mirror the existing
per-brand test coverage in `tests/testthat/`: palette identity against the
registry, theme element identity via `ggplot2::calc_element()`, scale output
via `ggplot2::ggplot_build()`, and (if a new font was bundled) that the
`.ttf` files ship and register offline.

## VHA in action

```{r vha-demo}

library(ggplot2)
library(scales)
library(hdatools)

housing_costs <- data.frame(
  locality = c("Locality A", "Locality B", "Locality C", "Locality D"),
  median_rent = c(1180, 1340, 1050, 1420)
)

ggplot(housing_costs, aes(median_rent, reorder(locality, median_rent), fill = locality)) +
  geom_col() +
  scale_fill_vha() +
  scale_x_continuous(labels = label_dollar()) +
  theme_vha(flip_gridlines = TRUE) +
  add_zero_line("x") +
  labs(
    title = "Median gross rent",
    subtitle = "Example localities (synthetic data)"
  )

```

A bare `theme_vha()` with no `scale_*()` call also brands the plot (ggplot2
>= 4.0's theme-carried palette):

```{r vha-theme-only}

ggplot(housing_costs, aes(median_rent, reorder(locality, median_rent), fill = locality)) +
  geom_col() +
  scale_x_continuous(labels = label_dollar()) +
  theme_vha(flip_gridlines = TRUE) +
  add_zero_line("x") +
  labs(
    title = "Median gross rent",
    subtitle = "No scale_fill_*() call — theme_vha() alone brands the fills"
  )

```
