---
title: "CVD accessibility audit — hdatools palettes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CVD accessibility audit — hdatools palettes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

This article documents a colorblind-vision (CVD) audit of the four hdatools brand
palettes conducted for the 0.4.0 release. It covers protanopia, deuteranopia, and
tritanopia at full severity. **No palette color order was changed as a result of this
audit** (see `plans/DECISIONS.md`, Q7, 2026-07-18). The findings here inform chart
authors on which color pairs to avoid in categorical plots and when supplemental
encodings are needed.

## Method

Pairwise perceptual distance (delta-E, CIE76) was computed in Lab color space via
`colorspace`:

1. Simulate full-severity CVD with `colorspace::protan()` / `deutan()` / `tritan()`
2. Convert simulated hex values to Lab with `colorspace::hex2RGB()` → `as(., "LAB")`
3. Compute Euclidean distance (delta-E) for every color pair

Delta-E below **10** is generally considered indistinguishable; values of **10–20** are
borderline; values above **20** are comfortably distinct. The audit focuses on the first
four palette slots — the range almost always used in a categorical plot — plus two
pairs flagged during the 0.4.0 design review.

```{r helpers, echo = FALSE}
library(colorspace)

hex_to_lab <- function(hex) {
  coords(methods::as(hex2RGB(hex), "LAB"))
}

pairwise_delta_e <- function(hex_vec) {
  nm <- names(hex_vec)
  n  <- length(hex_vec)
  rows <- vector("list", n * (n - 1L) / 2L)
  k <- 1L
  for (i in seq_len(n - 1L)) {
    for (j in seq(i + 1L, n)) {
      dE <- sqrt(sum((hex_to_lab(hex_vec[[i]]) - hex_to_lab(hex_vec[[j]]))^2))
      rows[[k]] <- data.frame(color1 = nm[i], color2 = nm[j],
                              delta_e = round(dE, 1),
                              stringsAsFactors = FALSE)
      k <- k + 1L
    }
  }
  do.call(rbind, rows)
}

cvd_table <- function(hex_vec) {
  sim_types <- c(Normal = "normal", Protan = "protan",
                 Deutan = "deutan", Tritan = "tritan")
  lapply(names(sim_types), function(label) {
    s <- sim_types[[label]]
    sim <- if (s == "normal") hex_vec else {
      fn <- switch(s, protan = protan, deutan = deutan, tritan = tritan)
      structure(fn(hex_vec, sev = 1), names = names(hex_vec))
    }
    df <- pairwise_delta_e(sim)
    df$cvd <- label
    df
  }) |> do.call(what = rbind)
}

show_table <- function(brand_name, hex_vec) {
  df  <- cvd_table(hex_vec)
  wide <- reshape(df,
    idvar     = c("color1", "color2"),
    timevar   = "cvd",
    direction = "wide",
    v.names   = "delta_e"
  )
  names(wide) <- sub("delta_e\\.", "", names(wide))
  wide$Flag <- ifelse(
    apply(wide[, c("Protan", "Deutan", "Tritan")], 1,
          function(r) any(r < 10, na.rm = TRUE)),
    "⚠️", ""
  )
  knitr::kable(wide, row.names = FALSE,
               caption = sprintf("%s palette — pairwise delta-E by CVD type", brand_name))
}
```

---

## HDA

Palette: Blue `#445ca9`, Green `#8baeaa`, Yellow `#e9ab3f`, Coral `#e76f52`,
Lavender `#a97a92`, Sea Green `#8abc8e`

```{r hda-table, echo = FALSE}
hda_pal <- c(
  "Blue"      = "#445ca9",
  "Green"     = "#8baeaa",
  "Yellow"    = "#e9ab3f",
  "Coral"     = "#e76f52",
  "Lavender"  = "#a97a92",
  "Sea Green" = "#8abc8e"
)
show_table("HDA", hda_pal[1:4])
```

**First-4 audit — no failures.** The minimum pairwise delta-E across all CVD types is
22.7 (Yellow vs Coral under deuteranopia). All first-4 pairs are comfortably distinct.

### HDA: Green vs Sea Green (flagged pair)

```{r hda-flagged, echo = FALSE}
show_table("HDA — Green vs Sea Green", hda_pal[c("Green", "Sea Green")])
```

**Tritanopia failure confirmed.** Green (`#8baeaa`) and Sea Green (`#8abc8e`) collapse to
delta-E = 5.97 under full tritanopia — effectively indistinguishable. Under protanopia
and deuteranopia the pair is acceptable (delta-E ≈ 20–22).

Tritanopia affects approximately 0.1 % of the population. Green and Sea Green are
positions 2 and 6, so they appear together only in a 5–6-category plot. Recommendations:

- **Avoid pairing Green and Sea Green in a plot that targets tritanopic accessibility.**
- In 2–4-category plots the pair will not both appear, so no action is needed.
- If a chart must use both and CVD robustness is critical, replace one with a manual
  override (e.g. swap Sea Green for Coral or Lavender).

---

## HFV

Palette: Shadow `#334a66`, Sky `#66cccc`, Lilac `#a29dd4`, Grass `#50aaa7`,
Berry `#c0327e`, Desert `#ec7c53`, Leaf `#6fb547`, Cerulean `#7fc7e0`

```{r hfv-table, echo = FALSE}
hfv_pal <- c(
  "Shadow"   = "#334a66",
  "Sky"      = "#66cccc",
  "Lilac"    = "#a29dd4",
  "Grass"    = "#50aaa7",
  "Berry"    = "#c0327e",
  "Desert"   = "#ec7c53",
  "Leaf"     = "#6fb547",
  "Cerulean" = "#7fc7e0"
)
show_table("HFV", hfv_pal[1:4])
```

**Sky vs Grass is the structurally risky pair.** Delta-E is approximately 12 under all
three CVD types (12.1–12.7) — within the borderline range and the lowest minimum of
any first-4 audit. Both colors are teal-family; they are distinguishable in normal
vision by their lightness contrast but that contrast diminishes under red-green CVD.

No pair falls below 10 in the first-4 audit, so no automatic failure. Recommendations:

- **Avoid using Sky and Grass as the only two series colors** in a CVD-sensitive chart.
  If both are needed, add a shape, pattern, or direct-label encoding alongside color.
- 3-category or 4-category plots should order series so Sky and Grass are not adjacent
  in a legend (where label proximity compensates for color proximity).

---

## PHA

Palette: Green `#5bab8e`, Light Blue `#a6cccc`, Orange `#f39152`, Red `#be451c`,
Purple `#a5add0`, Dark Blue `#2b6b9c`

```{r pha-table, echo = FALSE}
pha_pal <- c(
  "Green"      = "#5bab8e",
  "Light Blue" = "#a6cccc",
  "Orange"     = "#f39152",
  "Red"        = "#be451c",
  "Purple"     = "#a5add0",
  "Dark Blue"  = "#2b6b9c"
)
show_table("PHA", pha_pal[1:4])
```

**Green vs Light Blue is the tightest pair** (delta-E ≈ 18.5–21.3 across CVD types).
Both are cool desaturated colors that lose further differentiation under red-green CVD.
No pair falls below 10.

### PHA: Orange vs Red (flagged pair)

```{r pha-flagged, echo = FALSE}
show_table("PHA — Orange vs Red", pha_pal[c("Orange", "Red")])
```

**Not confirmed as a failure.** The flagged Orange/Red pair maintains delta-E ≥ 22 under
all CVD types; the lightness difference (Orange is substantially lighter) provides the
primary perceptual cue and it is preserved under CVD simulation.

Recommendation: no action needed for the Orange/Red pair. For Green vs Light Blue in
4-category plots, consider adding a secondary encoding if tritanopic accessibility is a
concern.

---

## VHA

Palette: Dark Turq `#0C4D4F`, Light Green `#A0D18E`, Yellow `#ECC51E`,
Light Turq `#19787B`, Grey `#2E3030`, Light Blue `#E3F3F5`

```{r vha-table, echo = FALSE}
vha_pal <- c(
  "Dark Turq"   = "#0C4D4F",
  "Light Green" = "#A0D18E",
  "Yellow"      = "#ECC51E",
  "Light Turq"  = "#19787B",
  "Grey"        = "#2E3030",
  "Light Blue"  = "#E3F3F5"
)
show_table("VHA", vha_pal[1:4])
```

**Dark Turq vs Light Turq is the tightest pair** (delta-E ≈ 16.5–18.3 across CVD
types). Both are deep teal tones; the primary cue under CVD is the lightness difference
(Dark Turq is considerably darker). No pair falls below 10 in the first-4 audit.

Recommendations:

- In 2-category plots the pair is acceptable; the lightness contrast is sufficient.
- In 4-category plots, prefer not to place Dark Turq and Light Turq as the only two
  categories without supplemental encoding.

---

## Diverging ramp caveat

As documented in NEWS.md, 7-class diverging maps built from any of the six continuous
ramps lose sign distinction in the innermost class pair under protanopia, because both
arms converge toward the shared cream center. This is structural to any cream-centered
diverging ramp; the mitigation is to always pair a `palette = "diverging"` map with a
legend or direct class labels.

---

## Regression tests

`tests/testthat/test-cvd.R` contains `colorspace::simulate_cvd()`-based assertions
that the first four colors of each brand palette maintain the minimum pairwise delta-E
thresholds observed above. These tests guard against future palette edits that silently
reduce CVD distinguishability; they do not validate the absolute CVD safety of the
palettes, and they do not test the HDA Green/Sea Green tritanopia failure (which is a
documented limitation, not a threshold a future edit could improve accidentally).
